Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization

### 2023-07-14 report/draft.Rmd
new version
minor textual changes

# Introduction
The behavior of citizens and the study of them by social scientists rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational]. 
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @grimmer2022text; @wei2023overview].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

We have conducted the survey experiments in the Netherlands in May 2022 and in the United States in January 2023.
Both samples, recruited through [KiesKompas](https://www.kieskompas.nl/en/) and [Prolific](https://www.prolific.co/) for respectively the Dutch and American case, consist of 3,000 participants (based on the power analysis presented in our [online compendium](LINK)) of 18 years and older.
Both survey companies works with non-random opt-in respondents.
Therefore, we measured many demographic background variables, and balance checks have been conducted to demonstrate whether certain categories are over represented in a certain experimental group.
Our study has been approved by the [Research Ethics Review Committee](https://fsw.vu.nl/nl/onderzoek/research-ethics-review/index.aspx) of the *Vrije Universiteit Amsterdam* (see the approval [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/docs/2022-3-30-59.pdf)).
To ensure good quality of our data, one attention check (discussed in more detail in [Section 3.3](#attention-checks)) is included [@berinsky2014separating].

@@

***Experimental Conditions.*** Respondents are randomly assigned to either view a political party as an actor, or a masked condition, where they see `X` as an actor; simultaneously, respondents see either a fully specified sentence or a underspecified sentence, in which one needs additional information to interpret the position on an actor.
Table \ref{tab:conditions} gives an overview of the variations in treatment in the surveys.

@@

# Discussion
Human annotation represents a crucial instrument in the toolkit of the social sciences.
In order to address concerns regarding the validity and reliability of annotation tasks, we establish strict standards.
Especially in these days, where due to the democratization of data and the advances in NLP more data can be analyzed or classified, these standards are more important than ever.
After all, an algorithm trained on biased data will reproduce and often exacerbate bias [e.g. see @prost2019debiasing].
It is key to realize that the underlying assumption to create valid and reliable annotated data is that there is only one correct interpretation for every input example, and that disagreement between the annotators is something that needs to be dealt with at all costs [@aroyo2015truth].
In times of increasing factual opinion polarization [@rekker2022understanding; @lee2021more], it is crucial to ask *who's perspective is being recorded in these datasets*, given that no neutral data and no apolitical standpoint from where we can call out bias: "Datasets are always 'a worldview' and, as such, data always remains biased" [@miceli2022studying, p.5].
Our tools to create valid and reliable annotation data does currently not allow for dealing with different perspectives. 
In this paper, we examined annotation bias in the classification of political stances using two pre-registered crowd-coding experiments in the Netherlands and the United States.
We used both a strict and a lenient interpretation of stance to illustrate whether decisions made in the research process on what constitutes a stance influences the results.

>>>>

# Introduction
The behavior of citizens and the study thereof by social scientists both rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational]. 
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @grimmer2022text; @wei2023overview].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

We have conducted the survey experiments in the Netherlands in May 2022 and in the United States in January 2023.
Both samples, recruited through [KiesKompas](https://www.kieskompas.nl/en/) and [Prolific](https://www.prolific.co/) for respectively the Dutch and American case, consist of 3,000 participants (based on the power analysis presented in our [online compendium](LINK)) of 18 years and older.
Both survey companies works with non-random opt-in respondents.
Therefore, we measured many demographic background variables, and balance checks have been conducted to demonstrate whether certain categories are over represented in a certain experimental group.
Our study has been approved by the [Research Ethics Review Committee](https://fsw.vu.nl/nl/onderzoek/research-ethics-review/index.aspx) of the *Vrije Universiteit Amsterdam* (see the approval [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/docs/2022-3-30-59.pdf)).
To ensure good quality of our data, one attention check (discussed in more detail in **OA XX** is included [@berinsky2014separating].

@@

***Experimental Conditions.*** Respondents are randomly assigned to either view a political party as an actor, or a masked condition, where they see `X` as an actor; simultaneously, respondents see either a fully specified sentence or a underspecified sentence, in which one needs additional information outside of the text to determine an actor's position.
Table \ref{tab:conditions} gives an overview of the variations in treatment in the surveys.

@@

# Discussion

Human annotation is a crucial instrument in the toolkit of the social sciences.
In order to safeguard the validity and reliability of annotation tasks, strict standards need to be upheld.
Especially in these days, where due to the democratization of data and the advances in NLP more data can be analyzed or classified, these standards are more important than ever.
After all, an algorithm trained on biased data will reproduce and often exacerbate bias [e.g. see @prost2019debiasing].
It is key to realize that the underlying assumption to create valid and reliable annotated data is that there is only one correct interpretation for every input example, and that disagreement between the annotators is something that needs to be dealt with at all costs [@aroyo2015truth].
In times of increasing factual opinion polarization [@rekker2022understanding; @lee2021more], it is crucial to ask *who's perspective is being recorded in these datasets*, given that no neutral data and no apolitical standpoint from where we can call out bias: "Datasets are always 'a worldview' and, as such, data always remains biased" [@miceli2022studying, p.5].
Our tools to create valid and reliable annotation data does currently not allow for dealing with different perspectives. 
In this paper, we examined annotation bias in the classification of political stances using two pre-registered crowd-coding experiments in the Netherlands and the United States.
We used both a strict and a lenient interpretation of stance to illustrate whether decisions made in the research process on what constitutes a stance influences the results.
